Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns

### 2019-07-23 ms/manuscript.Rmd
Add SI

---
title: "Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns"
bibliography: "references.bib"
mainfont: Times New Roman.ttf
fontsize: 12pt
mainfontoptions: 
- BoldFont=Times New Roman Bold.ttf
- ItalicFont=Times New Roman Italic.ttf
- BoldItalicFont=Times New Roman Bold Italic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
    includes:
      in_header: newphyt.sty
editor_options: 
  chunk_output_type: console
---

@@

# Load analysis results
loadd(climate_plot)
loadd(combined_comm_div_plots)
loadd(combined_cwm_plots)
loadd(comm)
loadd(correlated_evo_test)
loadd(correlated_evo_test)
loadd(diversity_data)
loadd(fern_traits)
loadd(model_summaries)
loadd(pca_plot)
loadd(pca_results)
loadd(phylosig_binary_traits)
loadd(phylosig_binary_traits)
loadd(phylosig_cont_traits)
loadd(pic_results)
loadd(t_test_results)
loadd(traits_with_tree)

```

@@

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} $\times$ \SI{10}{\meter} plots spanning an elevational gradient from \SIrange{20}{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} @[Nitta2017]. Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}). Each plot was divided into 25, \SI{2}{\meter} $\times$ \SI{2}{\meter} subplots. Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots). Epiphytic ferns were sampled by hand to \circa \SI{2}{\meter} on tree trunks or confirmed visually from the ground. Species were categorized as epiphytic or terrestrial based on field observations (see Trait Selection and Measurements below). The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities. Vouchers were deposited at GH and UC.

@@

Morphological traits of gametophytes were all qualitative (either categorical or binary; \tab \@ref(tab:traits-used)). Trait states were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017], grown from spores of known species in the lab, or taken from the literature. For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate. Non-cordate morphotypes include elongate (strap or ribbon \sensu @Farrar2008) and filamentous forms (\fig \@ref(fig:traits-used-image)). Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

@@

@@

We calculated mean daily maximum, mean, minimum, and standard variation of temperature, RH, and VPD for each site from the raw climate data, then grand means from the daily means. We analyzed correlation between the grand means with Pearson's correlation coefficient, and retained a set with correlation coefficients less than 0.9 (hereafter, "climatic variables"). We tested for differences between epiphytic and terrestrial dataloggers while accounting for elevation with ANCOVA, using growth habit as the categorical variable and elevation as the covariate for each climatic response variable. For visualizing the results, we selected the linear model with the lowest Akaike Information Criterion (AIC) out of the set of models including interaction effects and the effect of each independent variable alone. A single outlier site located on an exposed south-facing slope with unusually high temperature and low humidity (Mt. Rotui, 830 m) site was excluded from analysis.

For comparative analyses, we used the dated phylogenetic tree of [@Nitta2017], which includes all species from Moorea and some from the neighboring island of Tahiti. We trimmed the tree to species occurring only on Moorea ($\n = 130$ \si{\species}) using the \function{drop.tip} function of the \package{ape} package [@Paradis2004] in R [@RCoreTeam2015]. We tested for phylogenetic signal in quantitative traits using two metrics that relate the observed trait values to those expected under a model of trait evolution following Brownian motion (BM), Blomberg's *K* [@Blomberg2003] and Pagel's $\lambda$ [@Pagel1999], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012]. $\lambda$ is a scaling parameter that ranges from 0 (traits evolving randomly, as on a star phylogeny) to 1 (traits evolving along branches according to Brownian motion). \K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under a BM model of trait evolution: when $\K = 1$, traits are evolving according to BM; when $\K > 1$, traits are more conserved than expected under BM; when $\K < 1$, traits have less phylogenetic signal than expected under BM [@Blomberg2003]. We tested for significance in \K by comparing observed values of \K against a null distribution of \num{1000} trees with trait values shuffled randomly across the tips, and in $\lambda$ by conducting a likelihood ratio test comparing the log likelihoods of the observed value of $\lambda$ vs. $\lambda = 0$ (no phylogenetic signal). We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' *D* [@Fritz2010], which is designed to test for signal in binary traits. Values of \D range from 0 under a model of trait evolution by BM, to 1 for random distribution of traits; negative values (greater trait conservation than BM) and values greater than 1 (trait overdispersion) are possible. We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], which also conducts a significance test by comparing the observed value of \D with distributions of simulated values produced under one of two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM threshold model. The probability of obtaining the observed value of \D under either model is then calculated.

We used a multivariate approach to investigate differences in traits between epiphytes and terrestrial species. We used only quantitative traits, and included only species that had no missing data for any traits ($\n = 104$ \si{\species}). We first log-transformed traits as necessary to attain normality, then scaled each trait by its range. We subjected the transformed, scaled traits to a Principal Components Analysis (PCA) to reduce dimensionality. We conducted standard and phylogenetically corrected PCA, using the \function{PCA} and \function{phyl.pca} functions in the \package{FactoMineR} [@Le2008] and \package{phytools} R packages, respectively. The position of each species was plotted in the multi-dimensional trait space using the first two PC axes, then color-coded according to growth habit (epiphytic vs. terrestrial).

We detected significant phylogenetic signal in most traits, so we tested for differences between traits of epiphytic vs. terrestrial species using methods that can test for significant differences between traits associated with a binary character while accounting for phylogeny. For quantitative traits, we used the \function{brunch} function in the \package{caper} package [@Orme2013], which calculates phylogenetically independent contrasts in a quantitative character between alternative states of a binary trait. For qualitative (binary) traits, we used Pagel's test of correlated evolution [@Pagel1994] as implemented with the \function{fitPagel} function in the \package{phytools} package [@Revell2012]. This method compares the likelihood between two alternative models for a pair of binary traits. The first model tests the null hypothesis that rates of evolution of the two traits are independent. The second model tests the alternative hypothesis that the rates of evolution of one trait depends on the other. The likelihoods of obtaining the observed data are calculated for each model, then compared using a log likelihood test. A significantly better fit of the dependent model indicates that evolution of the two traits is correlated. To verify the effect of analysis method on our results, we performed alternative tests for correlation of growth habit with functional traits using phylogenetic generalized linear mixed models (PGLMMs; see Methods S1 and \tabs S1 and S2).

@@

Prior to testing for differences in diversity metrics, we first checked for correlation between metrics with Pearson's correlation coefficient, and retained a set of metrics with absolute correlation coefficients < 0.9. We then tested for differences between epiphytic and terrestrial communities with a paired t-test. We also conducted a more thorough analysis simultaneously testing the effects of environment and growth habit on diversity metrics using a full-subsets multiple regression approach with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018]. For each diversity metric, we built a set of models (Gaussian family, with site as a random effect) taking into account all pairwise interactions of explanatory variables, and ranked these by corrected AIC (AICc) to find the models that best explain the data. Our dataset included `r nrow(diversity_data) - nrow(grand_mean_climate_select)` communities that lacked climate data and could not be included in full-subsets multiple regression analysis. Since we found that all environmental variables are related to elevation (see Results), we also constructed linear models each diversity metric by growth habit with elevation as the covariate.

@@

```{r env-survey}
# loadd(climate_model_summaries)
# loadd(climate_models_parameter_summaries)
# 
# # Extract the slope of the minimum RH for epiphytes
# # (change in min. % rel humidity per meter).
# min_RH_epi_slope <- climate_models_parameter_summaries %>%
#   filter(var == "min_RH", term == "el") %>%
#   pull(estimate)
# 
# # Extract the slope of the minimum RH for terrestrial species.
# # In the results table, terrestrial slope is the difference
# # between the estimate for the elevation term and the
# # interaction term (add though because it's already negative).
# min_RH_ter_slope <-
#   magrittr::add(
#     min_RH_epi_slope,
#     climate_models_parameter_summaries %>%
#       filter(var == "min_RH", term == "el:habitterrestrial") %>%
#       pull(estimate)
# )

# There were significant interaction effects between growth habit and elevation on all microclimatic variables except for minimum temperature (Table S1). Slope and intercept for variables related to temperature were very similar between epiphytic and terrestrial dataloggers, but differed more strongly for those related to humidity (Table S1, Figure \@ref(fig:climate). Minimum RH increased `r min_RH_epi_slope %>% signif(2) %>% multiply_by(100) %>% percent()` per \SI{100}{\meter} elevation for epiphytic dataloggers, but only `r min_RH_ter_slope %>% signif(2) %>% multiply_by(100) %>% percent()` for terrestrial dataloggers (linear model, $\rval^2^ = \Sexpr{get_result(slopes.results, "r.squared", "min_RH")}$, $\pval = \Sexpr{get_result(slopes.results, "pval", "min_RH", TRUE)}$). The Mt. Rotui \SI{830}{\meter} plot was a clear outlier, with greater mean temperature and lower minimum RH relative to other plots at similar elevation (outliers in  \fig \ref{fig:climate} at \circa \SI{800}{\meter}). Mt. Rotui is a solitary peak located between Cook's Bay and Oponohu Bay. It is more isolated with greater exposure than the other two mountains where the rest of the plots were established, which are part of a mountain complex on the interior of the island (J. Nitta, pers. obs.).

```

Mean and standard deviation of temperature and relative humidity were retained after excluding variables with correlation coefficients >0.9.

@@

Terrestrial and epiphytic species are partly differentiated in trait space, occupying mostly distinct areas but overlapping in the middle regardless of method used (standard or phylogenetic PCA) (\fig \@ref(fig:pca)(c, d)). The first two PC axes explained `r percent(pca_std_cum_var)` and `r percent(pca_phy_cum_var)` of the variation in the trait data when analyzed using standard and phylogenetic PCA, respectively. The trait loadings were similar between the two analyses, except that most switched axes; for example, SLA was loaded on PC1 in the standard analysis but PC2 in the phylogenetic analysis (REF SUPP TABLE pca).

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs \@ref(tab:phylosig-cont), \@ref(tab:phylosig-binary), \fig \@ref(fig:tree)), and for quantitative traits, different results were obtained for $\lambda$ and K. When measured with $\lambda$, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of $\lambda$ close to 1; only number of pinna pairs had $\lambda$ close to zero (\tab \@ref(tab:phylosig-cont)). However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of $\K < 1$ (\tab \@ref(tab:phylosig-cont)). All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D = `r hairs_d`) or more conserved (gemmmae \D = `r gemmae_d`; glands, \D = `r glands_d`; morphotype, \D = `r morphotype_d`) than expected under a scenario of evolution by BM (\tab \@ref(tab:phylosig-binary)). Phylogenetic generalized linear mixed models (PGLMMs) also indicated significant signal in all gametophyte traits (REF SUPP TABLE). Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D = `r habit_d`; \tab \@ref(tab:phylosig-binary)).

@@

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (\tab \@ref(tab:pics)). Epiphytic species tend to have smaller values for all traits (\tab \@ref(tab:pics); contrasts were made by subtracting epiphytic from terrestrial values). Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, \pval = `r stipe_pval`), as well as frond width (\pval = `r width_pval`) and degree of laminae dissection (\pval = `r dissection_pval`). For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, \pval = `r morphotype_pval`; \tab \@ref(tab:corr-evo)). Quantitatively similar results were obtained using PGLMMs, except that no significant relationship was observed between rhizome diameter and growth habit, whereas SLA did show a significant relationship with growth habit (REF SUPP TABLE).

@@

```{r comm-div-numbers}
# Format model summaries for printing
model_summ_print <-
  model_summaries %>%
  mutate(r = round_t(r.squared, 2), 
         p = format_pval(p.value))

# Format t-test results for printing
t_test_results_print <-
  t_test_results %>%
  mutate(p = format_pval(p.value))

# Define simple function for extracting value of interest from
# diversity model results
get_mod_result <- function(resp_var_select, habit_select, 
                           indep_var_select, value) {
  model_summ_print %>% 
    filter(
      resp_var == resp_var_select, 
      habit == habit_select, 
      indep_var == indep_var_select) %>% 
    pull(!!enquo(value))
}

@@

# Calculate mean richness by growth habit
mean_rich <-
  diversity_data %>%
  group_by(habit) %>%
  summarize(
    mean_rich = mean(ntaxa) %>% round_t(1),
    sd_rich = sd(ntaxa) %>% round_t(1)
  ) %>%
  mutate(stat = glue("{mean_rich} ± {sd_rich}"))
```

<!-- NEED TO UPDATE TO REFLECT mean_RH instead of min_RH -->
On a single-trait basis, distribution of community-weighted mean values (CWMs) showed significant correlation with elevation for some traits, but no significant relationship was detected for any traits with minimum RH (\fig \@ref(fig:cwm-div)). 
Specific leaf area decreased with elevation for epiphytic (linear model, \rval^2^ = `r get_mod_result("sla", "epiphytic", "el", "r")`, \pval = `r get_mod_result("sla", "epiphytic", "el", "p")`) and terrestrial (linear model, \rval^2^ = `r get_mod_result("sla", "terrestrial", "el", "r")`, \pval = `r get_mod_result("sla", "terrestrial", "el", "p")`) communities. 
Degree of lamina dissection decreased with elevation in epiphytes (linear model, \rval^2^ = `r get_mod_result("dissection", "epiphytic", "el", "r")`, \pval = `r get_mod_result("dissection", "epiphytic", "el", "p")`), but not terrestrial communities. 
Stipe length showed a small, but significant, increase with elevation for epiphytic communities (linear model, \rval^2^ = `r get_mod_result("stipe", "epiphytic", "el", "r")`, \pval = `r get_mod_result("stipe", "epiphytic", "el", "p")`). 
Rhizome diameter decreased with elevation for terrestrial communities (linear model, \rval^2^ = `r get_mod_result("rhizome", "terrestrial", "el", "r")`, \pval = `r get_mod_result("rhizome", "terrestrial", "el", "p")`), but increased slightly in epiphytes (linear model, \rval^2^ = `r get_mod_result("rhizome", "epiphytic", "el", "r")`, \pval = `r get_mod_result("rhizome", "epiphytic", "el", "p")`).

We observed a total of `r traits_with_abun %>% filter(abundance > 0) %>% pull(species) %>% n_distinct` species across all plots, including `r n_epi` epiphytic and `r n_ter` terrestrial species. Mean species richness was `r mean_rich %>% filter(habit == "epiphytic") %>% pull(stat)` \si{\species}per plot for epiphytes and `r mean_rich %>% filter(habit == "terrestrial") %>% pull(stat)` \si{\species} per plot for terrestrial species (not significantly different; two-sided \tval-test, \pval = `r t_test_results_print %>% filter(resp_var == "ntaxa") %>% pull(p)`; \fig \@ref(fig:comm-div)). Species richness increased linearly with elevation and minimum RH for epiphytes, but not terrestrial communities (\fig \@ref(fig:comm-div)). Trait richness (FRic) was lower in epiphytic vs. terrestrial communities (two-sided \tval-test, \pval = `r t_test_results_print %>% filter(resp_var == "FRic") %>% pull(p)`), whereas functional diversity (FDiv) was lower in terrestrial communities (two-sided \tval-test, \pval = `r t_test_results_print %>% filter(resp_var == "FDiv") %>% pull(p)`), and functional evenness (FEve) did not differ between the two. FEve decreased with elevation, but not minimum RH, in terrestrial communities (\fig \@ref(fig:comm-div)). Epiphytic communities showed a trend of increasing trait richness along the environmental gradient, with greater FRic at high elevations (linear model, \rval^2^ = `r get_mod_result("FRic", "epiphytic", "el", "r")`, \pval = `r get_mod_result("FRic", "epiphytic", "el", "p")`) and plots with greater minimum RH (linear model, \rval^2^ = `r get_mod_result("FRic", "epiphytic", "mean_RH", "r")`, \pval = `r get_mod_result("FRic", "epiphytic", "mean_RH", "p")`). Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD (two-sided \tval-test, \pval = `r t_test_results_print %>% filter(resp_var == "mntd.obs.z") %>% pull(p)`) but not MPD (\pval = `r t_test_results_print %>% filter(resp_var == "mpd.obs.z") %>% pull(p)`; \fig \@ref(fig:comm-div)). No significant trend was detected in phylogenetic community structure with environment in either terrestrial or epiphytic communities (\fig \@ref(fig:comm-div)).

@@

The sporophyte traits that differed most strongly between epiphytic and terrestrial taxa were related to plant size, including stipe and frond length, frond width, and rhizome diameter. Stipe length in particular differed strongly between epiphytes and terrestrial species (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)). Stipe length may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance [@WatkinsJr.2010]. Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns [@WatkinsJr.2010; Creese2011]. However, given that many of these size-related traits scale together in ferns [@Arcand2008; Creese2011], careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

@@

\pagebreak

>>>>

---
title: "Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns"
bibliography: "references.bib"
mainfont: Times New Roman.ttf
fontsize: 12pt
mainfontoptions: 
- BoldFont=Times New Roman Bold.ttf
- ItalicFont=Times New Roman Italic.ttf
- BoldItalicFont=Times New Roman Bold Italic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
    includes:
      in_header: newphyt.sty
header-includes: 
  \usepackage{float}
  \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
editor_options: 
  chunk_output_type: console
---

@@

# Load analysis results
loadd(best_fit_div_models)
loadd(climate_ancova_results)
loadd(climate_plot)
loadd(climate_select)
loadd(combined_comm_div_plots)
loadd(combined_cwm_plots)
loadd(comm)
loadd(correlated_evo_test)
loadd(div_el_model_summaries)
loadd(div_t_test_results)
loadd(div_metrics_select)
loadd(fern_traits)
loadd(importance_heatmap)
loadd(pca_plot)
loadd(pca_results)
loadd(phylosig_binary_traits)
loadd(phylosig_cont_traits)
loadd(pic_results)
loadd(traits_with_tree)
```

@@

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} $\times$ \SI{10}{\meter} plots spanning an elevational gradient from \SIrange{20}{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017]. Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}). Each plot was divided into 25, \SI{2}{\meter} $\times$ \SI{2}{\meter} subplots. Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots). Epiphytic ferns were sampled by hand to \circa \SI{2}{\meter} on tree trunks or confirmed visually from the ground. Species were categorized as epiphytic or terrestrial based on field observations (see Trait Selection and Measurements below). The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities. Vouchers were deposited at GH and UC.

@@

Morphological traits of gametophytes were all qualitative (either categorical or binary; \tab \@ref(tab:traits-used)). Traits were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017] or grown from spores of known species in the lab, taken from the literature, or inferred from taxonomic group (i.e., genus) if documented as a stable taxonomic character. For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate. Non-cordate morphotypes include elongate (strap or ribbon \sensu @Farrar2008) and filamentous forms (\fig \@ref(fig:traits-used-image)). Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

@@

Trait values and data sources are summarized in Table S1.

@@

We calculated daily maximum, mean, minimum, and standard variation of temperature, RH, and VPD for each site from the raw climate data, then overall means from the daily values. We analyzed correlation between the mean values with Pearson's correlation coefficient, and retained a set with correlation coefficients less than 0.9 (hereafter, "climatic variables"). Variables relating to VPD, which have no meaning if less than zero, and some other variables that had skewed distributions, were square-root transformed prior to fitting models. We tested for differences between epiphytic and terrestrial dataloggers while accounting for elevation with ANCOVA, using growth habit as the categorical variable and elevation as the covariate. We identified the best-fitting linear model for each climatic variable in response to growth habit and elevation by constructing a set of models including the effect of each independent variable alone and their interactions, then selecting the one with the lowest Akaike Information Criterion (AIC). A single outlier site with unusually high temperature and low humidity (Mt. Rotui, 830 m) site was excluded from ANCOVA and linear models. Mt. Rotui is an isolated peak with greater exposure than the other two mountains where the rest of the plots were established, which are part of a mountain complex on the interior of the island.

For comparative analyses, we used the dated phylogenetic tree of [@Nitta2017], which includes all species from Moorea and some from the neighboring island of Tahiti. We trimmed the tree to species occurring only on Moorea ($\n = 130$ \si{\species}) using the \function{drop.tip} function of the \package{ape} package [@Paradis2004] in R [@RCoreTeam2015]. We tested for phylogenetic signal in quantitative traits using two metrics that relate the observed trait values to those expected under a model of trait evolution following Brownian motion (BM), Blomberg's *K* [@Blomberg2003] and Pagel's $\lambda$ [@Pagel1999], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012]. $\lambda$ is a scaling parameter that ranges from 0 (traits evolving randomly, as on a star phylogeny) to 1 (traits evolving along branches according to Brownian motion). \K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under a BM model of trait evolution: when $\K = 1$, traits are evolving according to BM; when $\K > 1$, traits are more conserved than expected under BM; when $\K < 1$, traits have less phylogenetic signal than expected under BM [@Blomberg2003]. We tested for significance in \K by comparing observed values of \K against a null distribution of \num{1000} trees with trait values shuffled randomly across the tips, and in $\lambda$ by conducting a likelihood ratio test comparing the log likelihoods of the observed value of $\lambda$ vs. $\lambda = 0$ (no phylogenetic signal). We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' *D* [@Fritz2010], which is designed to test for signal in binary traits. Values of \D range from 0 under a model of trait evolution by BM, to 1 for random distribution of traits; negative values (greater trait conservation than BM) and values greater than 1 (trait overdispersion) are possible. We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], which also conducts a significance test by comparing the observed value of \D with distributions of simulated values produced under one of two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM threshold model. To account for possible circularity in measuring the phylogenetic signal of traits that were scored based on taxonomy, we conducted the same analyses on a dataset excluding species that had been scored this way ("strict" dataset).

We used a multivariate approach to investigate differences in traits between epiphytes and terrestrial species. We used only quantitative traits, and included only species that had no missing data for any traits ($\n = 104$ \si{\species}). We first log-transformed traits as necessary to attain normality, then scaled each trait by its range. We subjected the transformed, scaled traits to a Principal Components Analysis (PCA) to reduce dimensionality. We conducted standard and phylogenetically corrected PCA, using the \function{PCA} and \function{phyl.pca} functions in the \package{FactoMineR} [@Le2008] and \package{phytools} R packages, respectively.

We detected significant phylogenetic signal in most traits, so we tested for differences between traits of epiphytic vs. terrestrial species using methods that can test for significant differences between traits associated with a binary character while accounting for phylogeny. For quantitative traits, we used the \function{brunch} function in the \package{caper} package [@Orme2013], which calculates phylogenetically independent contrasts in a quantitative character between alternative states of a binary trait. For qualitative (binary) traits, we used Pagel's test of correlated evolution [@Pagel1994] as implemented with the \function{fitPagel} function in the \package{phytools} package [@Revell2012]. This method compares the likelihood between two alternative models for a pair of binary traits. The first model tests the null hypothesis that rates of evolution of the two traits are independent. The second model tests the alternative hypothesis that the rates of evolution of one trait depends on the other. The likelihoods of obtaining the observed data are calculated for each model, then compared using a log likelihood test. A significantly better fit of the dependent model indicates that evolution of the two traits is correlated. The same analyses were also done on the "strict" dataset.

@@

Prior to testing for differences in diversity metrics, we first checked for correlation between metrics with Pearson's correlation coefficient, and retained a set of metrics with absolute correlation coefficients < 0.9. We tested for differences between epiphytic and terrestrial communities with a paired \tval-test. We simultaneously tested the effects of environment and growth habit on diversity metrics using a full-subsets multiple regression approach with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018]. For each diversity metric, we built a set of models (Gaussian family, with site as a random effect) taking into account all combinations of explanatory variables and their pairwise interactions, and ranked these by corrected AIC (AICc) to find the models that best explain the data. Our dataset included `r nrow(div_metrics_select) - nrow(climate_select)` communities that lacked climate data and could not be included in full-subsets multiple regression analysis. Since we found that all climatic variables are related to elevation (see Results), we also constructed linear models for each diversity metric using same model selection procedure as in the climate analysis.

@@

Mean, maximum, and standard deviation (SD) of temperature and mean, minimum, and SD of vapor pressure deficit (VPD) were retained after excluding variables with correlation coefficients > 0.9. Variables related to temperature changed with elevation but did not differ significantly between epiphytic and terrestrial dataloggers (\tab S2; \fig \ref{fig:climate}). Variables related to VPD differed by growth habit and elevation. Mean and maximum or minimum temperature and VPD all decreased with elevation, whereas SD of temperature increased, but SD of VPD decreased with elevation. Epiphytic dataloggers had higher values than terrestrial ones for variables related to VPD.

@@

Terrestrial and epiphytic species are partly differentiated in trait space, occupying mostly distinct areas but overlapping in the middle regardless of method used (standard or phylogenetic PCA) (\fig \@ref(fig:pca)(c, d)). The first two PC axes explained `r percent(pca_std_cum_var)` and `r percent(pca_phy_cum_var)` of the variation in the trait data when analyzed using standard and phylogenetic PCA, respectively. The trait loadings were similar between the two analyses, except that most switched axes; for example, SLA was loaded on PC1 in the standard analysis but PC2 in the phylogenetic analysis.

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs \@ref(tab:phylosig-cont), \@ref(tab:phylosig-binary), \fig \@ref(fig:tree)), and for quantitative traits, different results were obtained for $\lambda$ and K. When measured with $\lambda$, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of $\lambda$ close to 1; only number of pinna pairs had $\lambda$ close to zero (\tab \@ref(tab:phylosig-cont)). However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of $\K < 1$ (\tab \@ref(tab:phylosig-cont)). All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D = `r hairs_d`) or more conserved (gemmmae \D = `r gemmae_d`; glands, \D = `r glands_d`; morphotype, \D = `r morphotype_d`) than expected under a scenario of evolution by BM (\tab \@ref(tab:phylosig-binary)). Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D = `r habit_d`; \tab \@ref(tab:phylosig-binary)). Similar results were obtained using the "strict" dataset (\tab S3).

@@

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (\tab \@ref(tab:pics)). Epiphytic species tend to have smaller values for all traits (\tab \@ref(tab:pics); contrasts were made by subtracting epiphytic from terrestrial values). Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, \pval = `r stipe_pval`), as well as frond width (\pval = `r width_pval`) and degree of laminae dissection (\pval = `r dissection_pval`). For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, \pval = `r morphotype_pval`; \tab \@ref(tab:corr-evo)). Similar results were obtained using the "strict" dataset (\tab S4).

@@

```{r comm-div-numbers}
# Format model summaries for printing
model_summ_print <-
  div_el_model_summaries %>%
  mutate(r = round_t(r.squared, 2), 
         p = format_pval(p.value))

# Format t-test results for printing
t_test_print <-
  div_t_test_results %>%
  mutate(p = format_pval(p.value))

# Define simple function for extracting value of interest from
# diversity model results
get_mod_result <- function(data, resp_var_select, value) {
  data %>% 
    filter(var == resp_var_select) %>% 
    pull(!!enquo(value))
}

@@

# Calculate mean richness by growth habit
mean_rich <-
  div_metrics_select %>%
  group_by(habit) %>%
  summarize(
    mean_rich = mean(ntaxa) %>% round_t(1),
    sd_rich = sd(ntaxa) %>% round_t(1)
  ) %>%
  mutate(stat = glue("{mean_rich} ± {sd_rich}"))
```

In the full-subsets analysis including all climate variables and diversity metrics, growth habit and temperature emerged as the most important variables (\fig \ref{fig:heatmap}). Most of the the best-fitting models included growth habit or the interaction of growth habit and a temperature-related variable (\tab S5). Community-weighted means of frond length, frond width, stipe length, and rhizome diameter were correlated (Pearson's correlation coefficient > 0.9), so of these we only analyzed stipe length.

Distribution of community-weighted mean values (CWMs) differed between epiphytic and terrestrial communities overall (pinna number, stipe length) or co-varied with elevation (dissection, SLA) (\fig \@ref(fig:cwm-div)). Specific leaf area decreased with elevation, and was lower for epiphytic relative to terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "sla", "r")`, \pval = `r get_mod_result(model_summ_print, "sla", "p")`). Degree of lamina dissection decreased with elevation in epiphytes, but slightly increased in terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "dissection", "r")`, \pval = `r get_mod_result(model_summ_print, "dissection", "p")`). Stipe length and pinna number showed no significant variation with elevation, but were much lower overall for epiphytes relative to terrestrial communities (\tval-test, both \pval = `r get_mod_result(t_test_print, "stipe", "p")`).

We observed a total of `r traits_with_abun %>% filter(abundance > 0) %>% pull(species) %>% n_distinct` species across all plots, including `r n_epi` epiphytic and `r n_ter` terrestrial species. Mean species richness was `r mean_rich %>% filter(habit == "epiphytic") %>% pull(stat)` \si{\species}per plot for epiphytes and `r mean_rich %>% filter(habit == "terrestrial") %>% pull(stat)` \si{\species} per plot for terrestrial species (not significantly different; two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "ntaxa", "p")`; \fig \@ref(fig:comm-div)). Species richness increased  with elevation for both epiphytes and terrestrial communities (\fig \@ref(fig:comm-div)). Trait richness (FRic) was lower in epiphytic vs. terrestrial communities (two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "FRic", "p")`), whereas functional diversity (FDiv) was lower in terrestrial communities (two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "FDiv", "p")`), and functional evenness (FEve) did not differ between the two. Both epiphytic and terrestrial communities showed a trend of increasing trait richness along the environmental gradient, with greater FRic at high elevations (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "FRic", "r")`, \pval = `r get_mod_result(model_summ_print, "FRic", "p")`). Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD (two-sided \tval-test, \pval = `get_mod_result(t_test_print, "mntd.obs.z", "p")`) but not MPD (\pval = `r get_mod_result(t_test_print, "mpd.obs.z", "p")`; \fig \@ref(fig:comm-div)). No significant trend was detected in phylogenetic community structure with elevation in either terrestrial or epiphytic communities (\fig \@ref(fig:comm-div)).

@@

The sporophyte traits that differed most strongly between epiphytic and terrestrial taxa were related to plant size, including stipe and frond length, frond width, and rhizome diameter (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)). Stipe length may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance [@WatkinsJr.2010]. Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns [@WatkinsJr.2010; Creese2011]. However, given that these were all correlated with each other (Pearson's correlation coefficient > 0.9), and similar traits have been shown to scale together previously [@Arcand2008; Creese2011], careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

@@
